Review Paper Draft

### 2020-06-18 main.tex
Changing citation style; edits for 2.1 and 2.2

Relatedness describes how genetically similar individuals are, and can refer to both ancient and recent relatedness.  Recent relatedness is characterized by small clusters of high levels of similarity and is indicative of familial or cryptic relatedness. Ancient relatedness, or common ancestry, is generally characterized by large clusters of low levels of similarity. The presence of multiple distinct ancestry groups is known as population stratification.

Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. However, in this review we focus on methods to account for unobserved relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are be present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations. This phenomenon is commonly described as \textbf{confounding due to population stratification}. However, the mechanism of this phenomenon warrants some discussion. In particular, it is often overlooked that population structure itself does not confound the genotype-phenotype relationship in the classical sense, rather there must exist some non-genetic mechanism by which population stratification affects both the phenotype and genotype \cite{barton2019population, vilhjalmsson2012nature}. \\

@@

As an example, consider a genetic study to assess genetic variants associated with lung cancer in a sample comprised of subjects from two distinct subpopulationsk A and B. Assume the major allele of SNP $X$ is present with higher frequency in subpopulation A compared to subpopulation B, but it has no direct effect on lung cancer. Now suppose these subpopulations are geographically segregated in a such a way that subpopulation A is exposed to good air quality, and subpopulation B to poor air quality. A GWAS analysis of data from these subpopulations would find SNP $X$ to be significantly associated with lung cancer. This spurious association is not identified because of the presence of population stratification itself, but because that stratification is related to a causal environmental exposure. Indeed, if subpopulations A and B were not subject to different air qualities, all else being equal, SNP $X$ would not be found to be associated with the phenotype.\\

@@

The mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors influence population-specific allele frequencies, or vice versa, only that the two may be related. Additionally, we assume that the environmental exposure has no effect on on observed SNPs.

@@

An additional consideration in genetic studies that we will briefly mention is that of genetic background. Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \cite{vilhjalmsson2012nature} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. While we acknowledge the challenges presented by genetic background, this work will focus on correction methods for confounding due to environmental heterogeneity, for which genetic similarity may serve as a proxy.

>>>>

Relatedness describes how genetically similar individuals are, and can refer to both ancient and recent relatedness.  Recent relatedness is characterized by small clusters of high levels of similarity and is indicative of familial or cryptic relatedness. Ancient relatedness, or common ancestry, is generally characterized by large clusters of low levels of similarity. The presence of multiple distinct ancestry groups is known as population stratification. \pb{Should some of these terms, like cryptic relatedness, be bolded?}

Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. However, in this review we focus on methods to account for unobserved relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations. This phenomenon is commonly described as \textbf{confounding due to population stratification}. However, the mechanism of this phenomenon warrants some discussion. In particular, it is often overlooked that population structure itself does not confound the genotype-phenotype relationship in the classical sense, rather there must exist some non-genetic mechanism by which population stratification affects both the phenotype and genotype \cite{barton2019population, vilhjalmsson2012nature}. \\

@@

As an example, consider a genetic study to assess genetic variants associated with lung cancer in a sample comprised of subjects from two distinct subpopulations A and B. Assume the major allele of SNP $X$ is present with higher frequency in subpopulation A compared to subpopulation B, but it has no direct effect on lung cancer. Now suppose these subpopulations are geographically segregated in a such a way that subpopulation A is exposed to good air quality, and subpopulation B to poor air quality. Consequently, a GWAS analysis of data from these subpopulations would be predisposed to find SNP $X$ significantly associated with lung cancer. This spurious association is not identified because of the presence of population stratification itself, but because that stratification is related to a causal environmental exposure. Indeed, if subpopulations A and B were not subject to different air qualities, all else being equal, SNP $X$ would not be found to be associated with the phenotype.\\

@@

The mechanism of environmental confounding is summarized in the directed acyclic graph displayed in figure \ref{fig:ps_env}. The undirected dashed line between population stratification and environmental exposure indicates that while we assume these elements are correlated, we do not assume any causal direction. Otherwise stated, we do not assume that environmental factors influence population-specific allele frequencies, or vice versa, only that the two may be related. Additionally, we assume that the environmental exposure has no effect on on observed SNPs.  \pb{I find these previous two sentences confusing/contradictory: We assume that environment and genetics are related, but also that environment has no effect on SNPs.}

@@

An additional consideration in genetic studies that we will briefly mention is that of genetic background. Genetic background describes when SNPs that have a causal relationship with the phenotype of interest are not directly observed, but are correlated with SNPs that are observed. As \cite{vilhjalmsson2012nature} note, any trait that has some genetic basis will be subject to genetic background effects, just as traits that are influenced by the environment may be subject to environmental confounding. Given the importance of both genetic and environmental contributions to many complex phenotypes, both genetic background and environmental confounding are important factors in our ability to identify true causal genetic relationships between SNPs and phenotype. While we acknowledge the challenges presented by genetic background, this work will focus on correction methods for confounding due to environmental heterogeneity, for which genetic similarity may serve as a proxy.\pb{Overall comment about this paragraph: I feel like we *are* addressing genetic background, by conducting a genomewide multivariate analysis.  I think its worth pointing out that this is not the only way to deal with genetic background, but it's not as though we're not addressing it at all.}
